Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-23 docs/proposals/NSF/ProjDesc/ProjDesc.tex
went through entire proposal

An understanding of these filtration and contamination processes relies on the complex phenomenon of transport and dispersion in porous media \cite{scheidegger1961general, neuman1987stochastic, delgado2007longitudinal, zhang2016backward}. Further, the action of continually flowing groundwater can alter medium properties through erosive, transportive, and other processes. Thus, it becomes necessary to understand transport through complex media whose properties change {\em dynamically} in response to the intervening fluid flow. These effects are most noticeable during rapid events, like the gravitational collapse of a sinkhole~\cite{sandhu2018fate}, during which medium properties can change dramatically over short timescales. Though less obvious, the accumulation of slower process, such as mechanical or chemical erosion, can also substantially alter medium properties \cite{szymczak2009wormhole, ellis2013dissolution}.

{\bf This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow alters porous-media properties and to characterize the associated changes in dispersive transport within the evolving medium.} A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{neuman1987stochastic, moyner2016multiscale, lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep,topsep=0pt]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of fragments and grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/collage.pdf}
\caption{\label{collage} \em Geophysical examples of chemical and mechanical erosion. Top from left: Diagram of a coupled karst-soil system with conduit formation \cite{watson2019sinkholes}; Limestone pavement in Dent de Crolles, France; Nohoch Ch’en Sinkhole in Belize. Bottom: examples of soil erosion. The bottom right shows the microscopic soil texture, exhibiting polydispersity and anisotropy.}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

A major step towards the goal of developing high-fidelity simulations of dynamic porous-medium erosion is illustrated in Fig.~\ref{80circ8vel}. This figure shows simulated erosion of an initial configuration of 80 bodies of high volume fraction $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities such as contaminants.

@@

We first outline the theoretical foundation that will be used in later sections of the proposal. At the finest scale, the fluid flow intervening between individual constituents of a porous medium is governed by the incompressible Stokes equations
\begin{align}
  \label{eqn:StokesEq}
  \mu \Delta \uu = \grad p, \quad
  \grad \cdot \uu = 0, \quad \xx \in \Omega,
\end{align}
%\begin{equation}
%\label{eqn:StokesEq}
%\begin{split}
%\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
%\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
%\end{split}
%\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, $\mu$ the fluid viscosity, and $\Omega$ the fluid domain. Linear elliptic PDEs, such as the Stokes equations, can be reformulated as a boundary integral equation (BIE).  BIEs have several numerical advantages including a dimension reduction, automatic mass conservation, and high-order accuracy. An equivalent BIE formulation of the Stokes equations~\eqref{eqn:StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega,
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

Equation \eqref{eqn:StokesEq}, and its equivalent integral formulation, are microscopic descriptions required to model the shape evolution of individual fragments and grains due to fine-scale erosive effects, either mechanical or chemical. This microscopic description, however, becomes impractical when attempting to model seepage in very large porous systems where the microscopic geometry can be exceedingly complex. Instead, one must homogenize over microscopic details to arrive at an approximate description. Leading-order homogenization produces the well-known Darcy system 
\begin{align}
  \label{DarcyEq}
  \qq = - \bvec{K} \grad \phi
\end{align}
where $\qq$ is the {\em specific discharge} or {\em Darcy velocity}, and $\phi = z + p/(\rho g)$ is the hydraulic head, with $\rho$ and $g$ representing the fluid density and gravitational constant respectively.  Above, $\bvec{K}$ is the {\em conductivity}, which is related to the {\em permeability} $\bvec{k}$ through $\bvec{K} = \bvec{k} \rho g/\mu$.  Both the conductivity and permeability are in general tensors, and they reduce to scalars in the case of an isotropic medium. Note that the Darcy velocity is related to the tracer velocity through $\qq = \theta_f \uu$, where $\theta_f$ is porosity (i.e.~fluid volume fraction).  In the case of an isotropic medium, the Kozeny-Carman equation is often used to relate permeability to volume fraction~\cite{bear2013dynamics}
% See Bear Ch. 2.8, pg. 111
\begin{align}
  k = C \frac{\theta_f^3}{(1-\theta_f)^2}.
\end{align}
The macroscopic description offered by Eq.~\eqref{DarcyEq}, although approximate, simplifies analysis tremendously in that individual grains are no longer resolved and thus the domain geometry is generally much simpler.

@@

\paragraph{Broader Impacts:} The proposed research could lead to a range of societal benefits, including better management of water resources, accurate prediction of contaminant transport, and an understanding of the long-term effects of human activities, such as hydraulic fracturing or groundwater pumping, on porous media and water resources. In such scenarios, the slow timescales of mechanical and chemical erosion are coupled to the faster timescales of human-induced changes and gravitational collapse. Currently, the relationship between these scales is poorly understood. The proposed high-fidelity, efficient computational methods will enable a deeper understanding of these processes, which is the first step towards developing effective strategies for managing water resources and mitigating catastrophic events such as contamination or structural collapse. For example, the insights gained here could potentially help identify locations in natural aquifers vulnerable to contamination and/or collapse.

In addition to these societal impacts, the proposed work will spur the training of mathematicians and computational scientists in fundamentally cross-disciplinary research of great societal importance. Further, the PIs will engage in educational and outreach activities to impact future generations of mathematical scientists, with particular focus on women and underrepresented minorities (URM). Both PIs are already actively involved in graduate and undergraduate research, as well as outreach to K-12 students. Further, we have successfully recruited women and underrepresented minorities in high school, undergraduate, graduate, and postdoctoral research.

\paragraph{Intellectual Merit:} The proposed research introduces a host of new computational challenges and opportunities. First the range of scales is vast: spatial scales range from microscopic granular constituents to large geological aquifers; timescales range from that of a sudden sinkhole collapse to years required mechanical and chemical erosion. The systems are inherently multicomponent, with coupling between the fluid and solid phases. Although the governing PDEs are linear, the presence of moving boundaries introduces nonlinear feedback between geometry and flow. To tackle these challenges, the PIs will combine cutting-edge computational tools with reduced-order modeling. Mixed-scale, deep neural networks will be used to learn from the data generated by high-fidelity numerical simulations to parameterize coarse-grained models based on the multiphase framework. Additionally, controlled laboratory experiments will be used to guide and verify theory developed herein.

@@

  \item The proposed research is interdisciplinary by nature, with immediate applications in the fields of geophysics and environmental sciences. The PIs already have experience collaborating with scientists from other disciplines \cite{Caves2018, eastham2019multiphase, Bolles2019, cur-spe-hie-obr-goo-qua2018} through FSU's Geophysical Fluid Dynamics Institute, an interdisciplinary unit comprised of a diverse set of faculty from home departments in example Earth, Atmospheric, and Ocean Sciences, Geology, Physics, Civil Engineering, Environmental Engineering, Mechanical Engineering, Scientific Computing, Mathematics, and others.

@@

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion in the Stokes flow regime~\cite{Quaife2018}, the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of porous-media configurations formed by erosion~\cite{chi-moo-qua2019}.

%----------------------------------------------------------------------%
\section{Proposed Research} 
%----------------------------------------------------------------------%
The goal of the proposed research is to develop computational methods to better understand erosion processes. Specific research tasks will include:
\begin{enumerate}[noitemsep,label=(\arabic*)]
  \item Simulating chemical erosion, and transport, including sedimentation and compaction, of eroding bodies

@@

  \item Simulating the transport of finite Peclet number concentrations in eroded porous geometries.

@@

The development of accurate and efficient numerical methods to solve the Stokes equations enables high-fidelity simulations of the erosion process. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, which we have already begun to investigate~\cite{Quaife2018}, erosion result from the shear stress $\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss$, where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{eqn:StokesEq}. The motion of the interface is proportional to the absolute value of shear 
\begin{align}
  \Vn = \CE \, \abs{\tau}.
\end{align}
In the case of chemically-induced erosion, a solute concentration field $c$ undergoes advection and diffusion:
\begin{align}
  \label{Ceq}
  \pd{c}{t} + \uu \cdot \grad c = \Diff \Lap c.
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \Diff \pd{c}{\ny}.
\end{equation}
Both erosion processes have been investigated by the PIs and collaborators in the context of single bodies in high Reynolds-number flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the context of a porous media at low Reynolds-number flows, numerical methods have been developed by the PIs to simulate shear-induced erosion~\cite{Quaife2018, chi-moo-qua2019}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques.  The boundary integral solver is accelerated by the fast multipole method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $\mathcal{O}(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vort.pdf}
\caption{\label{80circ8vort} \em Simulation of 80 bodies eroding in Stokes flow under the action of shear stress. Flow is left to right. Color represents vorticity, which provides a convenient way to visualize local shear rates. 
}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

Figs.~\ref{80circ8vel} and~\ref{80circ8vort} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed barycentric quadrature scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables more densely packed configurations. The color in Fig.~\ref{80circ8vort} represents the vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity.  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. As erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. This channelization process is made even more apparent by viewing the velocity magnitude in Fig.~\ref{80circ8vel}.

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows how certain medium properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies disintegrate. Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation~\eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases in time. Note that, unlike the area plot, the vertical axis is logarithmic and so the straight line observed at early times indicates that the resistivity decreases exponentially.  The much more modest decrease of area fraction is not sufficient to account for this exponential rate. Hence the reshaping process, in particular the formation of channels, plays a pivotal role in the medium's macroscopic resistivity to flow.

Lastly, the right-most plot of Fig.~\ref{fig3} shows new, preliminary results on how medium anisotropy evolves during the reshaping process of erosion. Anisotropy is calculated as the ratio of vertical to horizontal resistivity, with the vertical resistivity computed by rotating the configuration of bodies by 90 degrees. This plot shows the anisotropy to initial increase with time. Thus, the initially isotropic state becomes erased as bodies streamline and form primarily horizontal channels. Remarkably, the medium reaches an anisotropy of over 10, meaning that erosion can lead to a configuration that resists flow 10 times more in the vertical direction than the horizontal direction.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
  \caption{\label{fig3} \em Left: The area fraction of solid bodies versus time. Time is normalized by the vanishing time $t_f$.  Middle: The resistivity of the porous matrix decrease much more rapidly than area fraction, indicating the reshaping process to be essential. Right: The anisotropy (ratio of vertical to horizontal resistivity) as it varies in time. Erosion can create a highly anisotropic medium.} \end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

{\bf An immediate goal of the proposed work is to extend the range of physical effects captured in these simulations.} A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature method~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing~\cite{chi-moo-qua2019}. The quadrature method is a non-intrusive modification of the trapezoid rule and results in an error bound that is independent of the grain configuration. However, the current implementation does not yet fully exploit FMM acceleration. The computational challenge is to develop FMM expansions for the velocity, vorticity, and deformation-tensor layer potentials. Since these layer potentials each have distinct kernels, we will utilize kernel-independent FMMs~\cite{yin-bir-zor2004, fon-dar2009}, and their HPC implementations (for example, see \url{http://pvfmm.org}). Other challenges to be investigated include preconditioning strategies to reduce the number of required GMRES iterations, and time adaptivity to enable stable simulations to user-specified tolerances. PI Quaife is an expert in developing such algorithms~\cite{qua-bir2015b, qua-bir2016, qua-bir2015a, qua-cou-dar2018}.

>>>>

An understanding of these filtration and contamination processes relies
on the complex phenomenon of transport and dispersion in porous media
\cite{scheidegger1961general, neuman1987stochastic,
delgado2007longitudinal, zhang2016backward}. Further, the action of
continually flowing groundwater can alter medium properties through
erosive, transportive, and other processes. Thus, it becomes necessary
to understand transport through complex media whose properties change
{\em dynamically} in response to the intervening fluid flow. These
effects are most noticeable during rapid events, like the gravitational
collapse of a sinkhole~\cite{sandhu2018fate}, during which medium
properties can change dramatically over short timescales. Though less
obvious, the accumulation of slower processes, such as mechanical or chemical erosion, can also substantially alter medium properties \cite{szymczak2009wormhole, ellis2013dissolution}.

{\bf This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow alters porous-media properties and to characterize the associated changes in dispersive transport within the evolving medium.} A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{neuman1987stochastic, moyner2016multiscale, lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep,topsep=0pt]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of fragments and grains due to intervening fluid flow and sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/collage.pdf}
\caption{\label{collage} \em Geophysical examples of chemical and mechanical erosion. Top from left: Diagram of a coupled karst-soil system with conduit formation \cite{watson2019sinkholes}; Limestone pavement in Dent de Crolles, France; Nohoch Che'en Sinkhole in Belize. Bottom: Examples of soil erosion. The bottom right shows the microscopic soil texture, exhibiting polydispersity and anisotropy.}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

A major step towards the goal of developing high-fidelity simulations of
dynamic porous-medium erosion is illustrated in Fig.~\ref{80circ8vel}.
This figure shows simulated erosion of 80 bodies with an initial high
volume fraction of 50\%. The color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities such as contaminants.

@@

We first outline the theoretical foundation that will be used in later sections of the proposal. At the finest scale, the fluid flow intervening between individual constituents of a porous medium is governed by the incompressible Stokes equations
\begin{align}
  \label{eqn:StokesEq}
  \mu \Delta \uu = \grad p, \quad
  \grad \cdot \uu = 0, \quad \xx \in \Omega,
\end{align}
%\begin{equation}
%\label{eqn:StokesEq}
%\begin{split}
%\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
%\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
%\end{split}
%\end{equation}
where $\uu$ is the velocity field, $p$ is the pressure field, $\mu$ is the fluid viscosity, and $\Omega$ is the fluid domain. Linear homogeneous elliptic PDEs, such as the Stokes equations, can be reformulated as a boundary integral equation (BIE).  BIEs have several numerical advantages including a dimension reduction, automatic mass conservation, and high-order accuracy. An equivalent BIE formulation of the Stokes equations~\eqref{eqn:StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega,
\end{align}
where $\ff$ is the Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

Equation~\eqref{eqn:StokesEq}, and its equivalent BIE formulation, are microscopic descriptions required to model the shape evolution of individual fragments and grains due to fine-scale erosive effects, either mechanical or chemical. This microscopic description, however, becomes impractical when attempting to model seepage in very large porous systems where the microscopic geometry can be exceedingly complex. Instead, one must homogenize over microscopic details to arrive at an approximate description. Leading-order homogenization produces the well-known Darcy system 
\begin{align}
  \label{DarcyEq}
  \qq = - \bvec{K} \grad \phi
\end{align}
where $\qq$ is the {\em specific discharge} or {\em Darcy velocity}, and $\phi = z + p/(\rho g)$ is the hydraulic head, with $\rho$ and $g$ representing the fluid density and gravitational constant respectively.  Above, $\bvec{K}$ is the {\em conductivity}, which is related to the {\em permeability} $\bvec{k}$ through $\bvec{K} = \bvec{k} \rho g/\mu$.  Both the conductivity and permeability are in general tensors, and they reduce to scalars in the case of an isotropic medium. Note that the Darcy velocity is related to the tracer velocity through $\qq = \theta_f \uu$, where $\theta_f$ is the porosity (i.e.~fluid volume fraction).  In the case of an isotropic medium, the Kozeny-Carman equation is often used to relate permeability to volume fraction~\cite{bear2013dynamics}
% See Bear Ch. 2.8, pg. 111
\begin{align}
  k = C \frac{\theta_f^3}{(1-\theta_f)^2}.
\end{align}
The macroscopic description offered by Eq.~\eqref{DarcyEq}, although approximate, simplifies analysis tremendously in that individual grains are no longer resolved and thus the domain geometry is generally much simpler.

@@

\paragraph{Broader Impacts:} The proposed research could lead to a range of societal benefits, including better management of water resources, accurate prediction of contaminant transport, and an understanding of the long-term effects of human activities, such as hydraulic fracturing or groundwater pumping, on porous media and water resources. In such scenarios, the slow timescales of mechanical and chemical erosion are coupled to the faster timescales of human-induced changes and gravitational collapse. Currently, the relationship between these scales is poorly understood. The proposed high-fidelity and efficient computational methods will enable a deeper understanding of these processes, which is the first step towards developing effective strategies for managing water resources and mitigating catastrophic events such as contamination or structural collapse. For example, the insights gained here will potentially help identify locations in natural aquifers vulnerable to contamination, collapse, or both.

In addition to these societal impacts, the proposed work will spur the training of mathematicians and computational scientists in fundamentally cross-disciplinary research of great societal importance. Further, the PIs will engage in educational and outreach activities to impact future generations of mathematical scientists, with particular focus on women and underrepresented minorities. Both PIs are already actively involved in graduate and undergraduate research, as well as outreach to K-12 students. Further, we have successfully recruited women and underrepresented minorities in high school, undergraduate, graduate, and postdoctoral research.

\paragraph{Intellectual Merit:} The proposed research introduces a host of new computational challenges and opportunities. First the range of scales is vast: spatial scales range from microscopic granular constituents to large geological aquifers; timescales range from that of a sudden sinkhole collapse to years required for certain mechanical and chemical erosion processes. The systems are inherently multicomponent, with coupling between the fluid and solid phases. Although the governing PDEs are linear, the presence of moving boundaries introduces nonlinear feedback between geometry and flow. To tackle these challenges, the PIs will combine cutting-edge computational tools with reduced-order modeling. Mixed-scale, deep neural networks will be used to learn from the data generated by high-fidelity numerical simulations and to parameterize coarse-grained models based on the multiphase framework. Additionally, controlled laboratory experiments will be used to guide and verify theory developed herein.

@@

  \item The proposed research is interdisciplinary by nature, with immediate applications in the fields of geophysics and environmental sciences. The PIs have experience collaborating with scientists from other disciplines \cite{Caves2018, eastham2019multiphase, Bolles2019, cur-spe-hie-obr-goo-qua2018} through FSU's Geophysical Fluid Dynamics Institute, an interdisciplinary unit comprised of a diverse set of faculty from home departments in example Earth, Atmospheric, and Ocean Sciences, Geology, Physics, Civil Engineering, Environmental Engineering, Mechanical Engineering, Scientific Computing, Mathematics, and others.

@@

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion in the Stokes flow regime~\cite{Quaife2018}, the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of densly packed porous-media configurations formed by erosion~\cite{chi-moo-qua2019}.

%----------------------------------------------------------------------%
\section{Proposed Research} 
%----------------------------------------------------------------------%
The goal of the proposed research is to develop computational methods to better understand erosion processes. Specific research tasks will include:
\begin{enumerate}[noitemsep,label=(\arabic*)]
  \item Simulating mechanical and chemical erosion, as well as transport, including sedimentation and compaction, of eroding bodies.

@@

  \item Simulating the transport of finite Peclet number concentrations in eroded porous medium.

@@

The development of accurate and efficient numerical methods to solve the Stokes equations enables high-fidelity simulations of erosion processes. We will consider erosion induced by either fluid-mechanical forces and by chemical reaction, i.e.~dissolution. In the former case, which we have already begun to investigate~\cite{Quaife2018}, erosion result from the shear stress $\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss$, where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{eqn:StokesEq}. The motion of the interface is proportional to the absolute value of shear 
\begin{align}
  \Vn = \CE \, \abs{\tau}.
\end{align}
In the case of chemically-induced erosion, a solute concentration field $c$ undergoes advection and diffusion:
\begin{align}
  \label{Ceq}
  \pd{c}{t} + \uu \cdot \grad c = \Diff \Lap c.
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \Diff \pd{c}{\ny}.
\end{equation}
Both erosion processes have been investigated by the PIs and collaborators in the context of single bodies in high Reynolds-number flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the context of a porous media at low Reynolds-number flows, numerical methods have been developed by the PIs to simulate shear-induced erosion~\cite{Quaife2018, chi-moo-qua2019}. These simulations combine highly accurate BIE methods with stable interface evolution techniques.  The BIE solver is accelerated by the Fast Multipole Method (FMM)~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $\mathcal{O}(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vort.pdf}
\caption{\label{80circ8vort} \em Simulation of 80 bodies mechanically eroding in Stokes flow. Flow is left to right. Color represents vorticity, which provides a convenient way to visualize local shear rates. 
}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

Figs.~\ref{80circ8vel} and~\ref{80circ8vort} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric quadrature scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables more densely packed configurations. The color in Fig.~\ref{80circ8vort} represents the vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity.  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal---corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. As erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger transmit more flow, promoting local erosion rates and further widening these select channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. This channelization process is most evident by viewing the velocity magnitude in Fig.~\ref{80circ8vel}.

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows how certain medium properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies disintegrate. Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation~\eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases. Note that, unlike the area plot, the vertical axis is logarithmic, so the straight line observed at early times indicates that the resistivity decreases exponentially.  The much more modest decrease of area fraction is not sufficient to account for this exponential rate. Hence the reshaping process, in particular the formation of channels, plays a pivotal role in the medium's macroscopic resistivity to flow. Lastly, the right-most plot of Fig.~\ref{fig3} shows new, preliminary results on how medium anisotropy evolves during the reshaping process of erosion. Anisotropy is calculated as the ratio of vertical to horizontal resistivity, with the vertical resistivity computed by rotating the configuration of bodies by 90 degrees. This plot shows that the anisotropy initially increase with time. Thus, the initial isotropy is erased as bodies streamline and form primarily horizontal channels. Remarkably, the medium reaches an anisotropy of over 10, meaning that erosion can lead to a configuration that resists flow 10 times more in the vertical direction than the horizontal direction.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
  \caption{\label{fig3} \em Left: The area fraction of solid bodies versus time. Time is normalized by the vanishing time $t_f$.  Middle: The resistivity of the porous matrix decrease much more rapidly than the area fraction, indicating the reshaping process to be essential. Right: The anisotropy (ratio of vertical to horizontal resistivity) as it varies in time. Erosion can create a highly anisotropic medium.} 
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

{\bf An immediate goal of the proposed work is to extend the range of
physical effects captured in these simulations.} A particular challenge
of a BIE formulation is developing quadrature for nearly-singular
integrands that arise when bodies are close. To address this problem, we
have very recently extended a Barycentric quadrature
method~\cite{bar2014, bar-wu-vee2015} and have successfully simulated
erosion with grains that are initially separated by 1\% of an arclength
spacing~\cite{chi-moo-qua2019}. The quadrature method is a non-intrusive
modification of the trapezoid rule and results in an error bound that is
independent of the grain configuration. However, the current
implementation does not yet fully exploit FMM acceleration. The
computational challenge is to develop FMM expansions for the velocity,
vorticity, pressure, and deformation-tensor layer potentials. Since these layer potentials each have distinct kernels, we will utilize kernel-independent FMMs~\cite{yin-bir-zor2004, fon-dar2009}, and their HPC implementations (for example, see \url{http://pvfmm.org}). Other challenges to be investigated include preconditioning strategies to reduce the number of required GMRES iterations, and time adaptivity to enable stable simulations to user-specified tolerances. PI Quaife is an expert in developing such algorithms~\cite{qua-bir2015b, qua-bir2016, qua-bir2015a, qua-cou-dar2018}.
